Parsing Postfix log files

### 2009-02-28 doc/logparser-acronyms.tex
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

\newglossaryentry{SMTP}{
    type=\acronymtype,
    name={SMTP},
    description={Simple Mail Transfer Protocol},
    first={Simple Mail Transfer Protocol (SMTP)}
}
\newcommand{\SMTPglossaryDescription}[0]{%
    \acronym{SMTP} is the protocol used for transferring mail between the
    sending and receiving \acronym{MTA}\@.  It is a simple, human readable,
    plain text protocol, making it quite simple to test and debug problems
    with it.  A detailed description of SMTP is beyond the scope of this
    thesis: the original protocol definition is in RFC~821~\cite{RFC821},
    updated in RFC~2821~\cite{RFC2821}; introductory guides can be found
    at~\cite{smtp-intro-01,smtp-intro-02}.%
}
\newglossaryentry{SMTP glossary}{
    name={Simple Mail Transfer Protocol},
    description={
        \SMTPglossaryDescription{}
    }
}

>>>>

\newglossaryentry{SMTP}{
    type=\acronymtype,
    name={SMTP},
    description={Simple Mail Transfer Protocol},
    first={Simple Mail Transfer Protocol (SMTP)}
}
\newcommand{\SMTPglossaryDescription}[0]{%
    \acronym{SMTP} is the protocol used for transferring mail between the
    sending and receiving \acronym{MTA}\@.  It is a simple, human readable,
    plain text protocol, making it quite simple to test and debug problems
    with it.  A detailed description of SMTP is beyond the scope of this
    thesis: the original protocol definition is in RFC~821~\cite{RFC821},
    updated in RFC~2821~\cite{RFC2821}; introductory guides can be found at
    \urlLastChecked{http://db.glug-bom.org/lug-authors/philip/docs/mail-stuff/smtp-intro.html}{2009/02/28}
    if more information is needed.%
}
\newglossaryentry{SMTP glossary}{
    name={Simple Mail Transfer Protocol},
    description={
        \SMTPglossaryDescription{}
    }
}

### 2009-02-28 doc/logparser-architecture.tex
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

@@

Decoupling the rules from the actions allows external tools to be written
to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \acronym{FA} and can be converted to \acronym{FA},
so regex overlap can be detected by finding a non-empty intersection of two
\acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
for example in~\cite{intersection-of-NFA-using-Z}) uses DeMorgan's law:
$FA1~\cap{}~FA2~=~\overline{(\overline{FA1}~\cup{}~\overline{FA2})}$, which
has considerable computation complexity.  XXX COMMENT ON THE COMPUTATIONAL
COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE SPINNING WHEEL
BOOK\@.  Perl 5.10 regexes are more powerful than traditional regexes: it
is possible to match correctly balanced brackets nested to an arbitrary
depth, e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA} or a \acronym{CFL};
detecting overlap may require calculating the intersection of two
\acronym{PDA} or \acronyms{CFL} instead of two \acronym{FA}.  The
intersection of two \acronyms{CFL} is not closed, i.e.\ the resulting
language cannot always be parsed by a \acronym{CFL}, so intersection may be
intractable in some cases, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
complexity of calculating \acronym{FA} or \acronym{PDA} intersection.  This
is certainly not a task to be performed every time the parser runs: naive
detection of overlap amongst the Postfix log parser's \numberOFrules{}
rules would require calculating \numberOFruleINTERSECTIONS{} intersections.
When detecting overlap any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached which prevent the overlap in practice.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlapping regexes amongst the rules in each set
of rules with overlapping conditions.  Depending on the conditions employed
a rule may fall into more than one group, e.g.\ given these three
conditions:

>>>>

XXX INTEGRATE THE NEXT PARAGRAPH INTO THIS SECTION\@.

The solution developed is conceptually simple: provide a few generic
functions (\textit{actions\/}), each capable of dealing with an entire
category of inputs (e.g.\ rejecting a mail delivery attempt), accompanied
by a multitude of precise patterns (\textit{rules\/}), each of which
matches all inputs of a specific type and only that type (e.g.\ rejection
by a specific \acronym{DNSBL}).  It is an accepted standard to separate the
parsing procedure from the declarative grammar it operates with; part of
the novelty here is in the way that the grammar is itself partially
procedural (each action is a separate procedure).  This architecture is
ideally suited to parsing inputs where the input is not fully understood or
does not conform to a fixed grammar: the architecture warns about unparsed
inputs and other errors, but continues parsing as best it can, allowing the
developer of a new parser to decide which deficiencies are most important
and require attention first, rather than being forced to fix the first
error that arises.

@@

Decoupling the rules from the actions allows external tools to be written
to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \acronym{FA} and can be converted to \acronym{FA},
so regex overlap can be detected by finding a non-empty intersection of two
\acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
for example in~\cite{intersection-of-NFA-using-Z}) uses DeMorgan's law:
$FA1~\cap{}~FA2~=~\overline{(\overline{FA1}~\cup{}~\overline{FA2})}$, which
has considerable computation complexity.  XXX COMMENT ON THE COMPUTATIONAL
COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE SPINNING WHEEL
BOOK\@.  Perl 5.10 regexes~\cite{perlre} are more powerful than traditional
regexes: it is possible to match correctly balanced brackets nested to an
arbitrary depth, e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!
matches \verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires
the regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA} or a \acronym{CFL};
detecting overlap may require calculating the intersection of two
\acronym{PDA} or \acronyms{CFL} instead of two \acronym{FA}.  The
intersection of two \acronyms{CFL} is not closed, i.e.\ the resulting
language cannot always be parsed by a \acronym{CFL}, so intersection may be
intractable in some cases, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
complexity of calculating \acronym{FA} or \acronym{PDA} intersection.  This
is certainly not a task to be performed every time the parser runs: naive
detection of overlap amongst the Postfix log parser's \numberOFrules{}
rules would require calculating \numberOFruleINTERSECTIONS{} intersections.
When detecting overlap any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached which prevent the overlap in practice.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlapping regexes amongst the rules in each set
of rules with overlapping conditions.  Depending on the conditions employed
a rule may fall into more than one group, e.g.\ given these three
conditions:

### 2009-02-28 doc/logparser-glossary.tex
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

% vim: set filetype=tex :
% The contents of the glossary.
% XXX CHECK FOR ENTRIES THAT SHOULD BE INCLUDED.
\newglossaryentry{SQLite}{name={SQLite},description={
    \textit{SQLite is a small C library that implements a self-contained,
    embeddable, zero-configuration SQL database engine.\/} SQLite is an
    \acronym{SQL} implementation focusing on correctness, simplicity, and
    speed.  Unlike other \acronym{SQL} implementations it does not require
    a separate server process, greatly simplifying deployment of an
    application utilising it.  More details can be found
    at~\urlLastChecked{http://www.sqlite.org/}{2009/02/23}.
}}

@@

%\newglossaryentry{hash}{name={hash},description={
%    A hashing function transforms a string of characters to a number.
%    There are many possible uses for the resulting number: a common usage
%    is to maintain a data structure indexed by strings in an efficient
%    manner.  A full description is beyond the scope of this paper, further
%    information can be found at~\cite{hash-functions}.
%}}

@@

\newglossaryentry{mail bomb}{name={Mail Bomb},description={
    A mail bomb occurs when an attacker sends a huge volume of mail to a
    victim.  At best a mail bomb is irritating to the victim; at worst the
    deluge of mail can be severe enough to interrupt service for the victim
    and/or other users, by causing mail to be delayed or rejected because
    the victim or the server has reached a limit, e.g.\ too many mails in
    their mailbox, too much disk space consumed.
}}

\newglossaryentry{mail loop}{name={Mail Loop},description={
    Sometimes mail sent to one address must be forwarded to a different
    address instead, e.g.\ because a person has changed jobs.  A mail loop
    occurs when the chain of recipient addresses constitute a cyclic
    directed graph; the simplest example is when mail for
    \texttt{alice@example.net} is delivered to \texttt{bob@example.com},
    and mail for \texttt{bob@example.com} is delivered to
    \texttt{alice@example.net}.  XXX EXPLAIN WHY THIS IS NOT AUTOMATICLLY
    DETECTED SOMETIMES\@.
}}

>>>>

% vim: set filetype=tex :
% The contents of the glossary.
\newglossaryentry{SQLite}{name={SQLite},description={
    \textit{SQLite is a small C library that implements a self-contained,
    embeddable, zero-configuration SQL database engine.\/} SQLite is an
    \acronym{SQL} implementation focusing on correctness, simplicity, and
    speed.  Unlike other \acronym{SQL} implementations it does not require
    a separate server process, greatly simplifying deployment of an
    application utilising it.  More details can be found
    at~\urlLastChecked{http://www.sqlite.org/}{2009/02/23}.
}}

@@

\newglossaryentry{hash}{name={hash},description={
    A hashing function transforms a string of characters to a number.
    There are many possible uses for the resulting number: a common usage
    is to maintain a data structure indexed by strings in an efficient
    manner.  A full description is beyond the scope of this paper, further
    information can be found at~\cite{hash-functions}.
}}

@@

\newglossaryentry{mail bomb}{name={Mail Bomb},description={
    A mail bomb occurs when an attacker sends a huge volume of mail to a
    victim.  At best a mail bomb is irritating to the victim; at worst the
    deluge of mail can be severe enough to interrupt service for the victim
    and/or other users, by causing mail to be delayed or rejected because
    the victim or the server has reached a limit, e.g.\ too many mails in
    their mailbox, or too much disk space consumed by the mail queue.
}}

\newglossaryentry{mail loop}{name={Mail Loop},description={
    Sometimes mail sent to one address must be forwarded to a different
    address instead, e.g.\ because a person has changed jobs.  A mail loop
    occurs when the chain of recipient addresses constitute a cyclic
    directed graph; the simplest example is when mail for
    \texttt{alice@example.net} is delivered to \texttt{bob@example.com},
    and mail for \texttt{bob@example.com} is delivered to
    \texttt{alice@example.net}.  Postfix adds a \texttt{Delivered-To:}
    header to all mails it delivers locally to detect mail loops, but in
    some cases this header is removed by the program that forwards the
    mail.
}}

### 2009-02-28 doc/logparser-implementation.tex
Corrections to the background chapter.
Checks for lines containing only spaces and "log line" in the architecture chapter.
Corrections, improvements, and new content in architecture chapter. Added SGAI and Elsevier entries to bibliography. Some minor cleanup elsewhere.
Corrections from proofreading of architecture chapter.
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

@@

\label{overlapping rules in architecture}

XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?

XXX DFA COMPARISON\@.

@@

        In the original algorithm the purpose of the preprocessing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash~\cite{hash-functions} from a small range
        of values for each word of each log line, incrementing a counter
        for each hash.  The counters will be used in the next pass to
        filter out words: if the word's hash does not have a high
        frequency, the word itself cannot have a high frequency, so there
        is no need to maintain a counter for it, reducing the number of
        counters required and thus the program's memory consumption.

>>>>

Each log line contains the name of the Postfix component that produced it,
and this information is used when determining which rules should be used to
parse each log line (see \sectionref{rules in implementation} for details).

@@

\label{overlapping rules in implementation}

@@

        In the original algorithm the purpose of the preprocessing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash~\cite{hash-functions}\glsadd{hash} from a
        small range of values for each word of each log line, incrementing
        a counter for each hash.  The counters will be used in the next
        pass to filter out words: if the word's hash does not have a high
        frequency, the word itself cannot have a high frequency, so there
        is no need to maintain a counter for it, reducing the number of
        counters required and thus the program's memory consumption.

### 2009-02-28 doc/logparser-introduction.tex
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam measures are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the author's first real foray into processing
Postfix log files.}  Very few will have performed in-depth parsing and
analysis of their log files, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this kind of processing is the
unstructured nature of Postfix log files, where each log line was added on
an ad hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from
\urlLastChecked{http://www.postfix.org/}{2009/02/23}} Further complication
arises because the set of rejection messages is not fixed: new messages can
be added by the administrator with custom checks; every
\acronym{DNSBL}\footnote{This document is supplied with a glossary, see
\textsection\ref{Glossary}.} returns a different explanatory message;
policy servers may log different messages depending on the characteristics
of the connection.  There are many ways in which the log lines may differ
between servers, even within the same organisation: servers may be
configured differently, or running different versions of Postfix.  This
paper documents the difficult process of parsing Postfix log files,
presenting \parsername{}, a program that parses Postfix log files and
places the resulting data into a database for later use.  The gathered data
can then be used to optimise current anti-spam measures, provide a baseline
to test new anti-spam measures against, or to produce statistics showing
how effective those measures are.  Numerous other uses are possible for
such data: improving server performance by identifying troublesome
destinations and reconfiguring appropriately; identifying regular high
volume uses (e.g.\ customer newsletters) and restricting those uses to
off-peak times; detecting virus outbreaks that propagate via mail; as a
base for billing customers on a shared server.  Preserving the raw data
enables users to develop a multitude of uses far beyond those conceived of
by the author.

@@

@@

Chapter~\ref{background} provides background information useful in
understanding the thesis, parser, and architecture.

Chapter~\ref{state of the art review} reviews both the previously published
research in this area and other available Postfix log file parsers,
discussing why they were deemed unsuitable for the task, including why they
could not be improved or expanded upon.

@@

\section{XXX NOTES}

Portions of \sectionref{parser architecture}, \sectionref{Postfix Parser
Implementation}, and \sectionref{Results} have previously been published at
a conference~\cite{sgai-2008}, and later reprinted in a
journal~\cite{elsevier-2009}.  Publication of the conference paper and
journal article above was supported by Science Foundation Ireland
RFP~05/RF/CMS002.

>>>>

The architecture described herein was developed as part of a larger project
to improve anti-spam defences by analysing the performance of the set of
filters currently in use, optimising the order and membership of the set
based on that analysis, and developing supplemental filters where
deficiencies are identified.  Most anti-spam techniques are content-based
(e.g.~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}) and
require the mail to be accepted before determining if it is spam, but
rejecting mail during the delivery attempt is preferable: senders of
non-spam mail that is mistakenly rejected will receive an immediate
non-delivery notice; resource usage is reduced on the accepting mail server
(allowing more intensive content-based techniques to be used on the mail
that is accepted); users have less spam mail to wade through.  Improving
the performance of anti-spam techniques that are applied when mail is being
transferred via \acronym{SMTP} is the goal of this project, by providing a
platform for reasoning about anti-spam filters.  The approach chosen to
measure performance is to analyse the log files produced by the
\acronym{SMTP} server in use, Postfix
(\urlLastCheckedNoParens{http://www.postfix.org/}{2009/02/28}), rather than
modifying its source code to generate statistics: this approach improves
the chances of other Postfix users testing and using the software developed
for this project.  The need arose for a parser capable of dealing with the
great number and variety of log lines produced by Postfix: the parser must
be designed so that adding support for parsing new inputs is a simple task,
because the log lines to be parsed will change over time.

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam measures are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the author's first real foray into processing
Postfix log files.}  Very few will have performed in-depth parsing and
analysis of their log files, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this kind of processing is the
unstructured nature of Postfix log files, where each log line was added on
an ad hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from
\urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\acronym{DNSBL}\footnote{This thesis is supplied with a glossary
(\textsection\ref{Glossary}) and a list of acronyms
(\textsection\ref{Acronyms}).} returns a different explanatory message;
policy servers may log different messages depending on the characteristics
of the connection.  There are many ways in which the log lines may differ
between servers, even within the same organisation: servers may be
configured differently, or running different versions of Postfix.

This thesis documents the difficult process of parsing Postfix log files,
presenting an architecture designed to enable the users of a parser to
easily extend it to parse their particular inputs, without requiring a lot
of work or a high level of understanding of the parsing process and the
parser's internal workings.  This architecture is the basis of
\parsername{}, a program that parses Postfix log files and places the
resulting data into a database for later analysis.  The gathered data can
be used to optimise current anti-spam measures, provide a baseline to test
new anti-spam measures against, or to produce statistics showing how
effective those measures are.  Numerous other uses are possible for such
data: improving server performance by identifying troublesome destinations
and reconfiguring appropriately; identifying regular high volume uses
(e.g.\ customer newsletters) and restricting those uses to off-peak times
or providing a dedicated service for them; detecting virus outbreaks that
propagate via mail; as a base for billing customers on a shared server.
Preserving the raw data enables users to develop a multitude of uses far
beyond those conceived of by the author.

@@

Chapter~\ref{background} provides background information useful in
understanding the motivation behind the project, \acronym{SMTP}, and
Postfix.

Chapter~\ref{state of the art review} reviews both the previously published
research in this area and other Postfix log file parsers, discussing why
they were deemed unsuitable for the task, including why they could not be
improved or expanded upon.

@@

@@

\section{Previously Published Work}

Portions of \textsection\ref{introduction}, \textsection\ref{parser
architecture}, \textsection\ref{Postfix Parser Implementation}, and
\textsection\ref{Results} have previously been published at a
conference~\cite{sgai-2008}, and later reprinted in a
journal~\cite{elsevier-2009}.  Publication of the conference paper and
journal article was supported by Science Foundation Ireland
RFP~05/RF/CMS002.

### 2009-02-28 doc/logparser-other-parsers-review.tex
Corrections, improvements, and new content in architecture chapter. Added SGAI and Elsevier entries to bibliography. Some minor cleanup elsewhere.
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

At the start of this project ten Postfix log file parsers were tested, with
the hope of finding a suitable parser to build upon, rather than starting
from scratch.  It was quite difficult to find ten parsers to review for
this project, and the functionality offered by those parsers ranges from
quite basic to much more mature, depending on the needs of the author of
the parser.  None of those parsers were suitable for this project, so the
decision was taken to write a new parser.  The first parser reviewed is the
only previously published research in this area that the author is aware
of; it parses Postfix log files, but aim of the research is to show that
presenting extracted data in a more easily accessible format is useful to
systems administrators, rather than to improve anti-spam techniques.

The same ten parsers have been reviewed and compared with \parsername{},
this project's finished parser, to show how much effort would have been
required to fulfil the aims and requirements of this project.  It is
important to compare and contrast newly developed algorithms and parsers
against those already available, to accurately judge what improvements, if
any, are delivered by the newcomers.  Each of the reviewed parsers was
tested with the \numberOFlogFILES{} test log files described in
\sectionref{parser efficiency}.  \Tableref{Summary of parsers' features}
summarises the results of this review.

@@

>>>>

At the start of this project ten Postfix log file parsers were tested, with
the hope of finding a suitable parser to build upon, rather than starting
from scratch.  It was quite difficult to find ten parsers to review for
this project, and the functionality offered by those parsers ranges from
quite basic to much more mature, depending on the needs of the author of
the parser.

It was hoped to reuse an existing parser rather than writing one from
scratch, but the existing parsers considered were rejected for one or more
reasons.  The effort required to adapt and improve an existing parser was
judged to be greater than the effort to write a new one, because the
techniques used by the existing parsers severely limited their potential:
some ignored the majority of log lines, parsing specific log lines
accurately, but without any provision for parsing new or similar log lines;
others sloppily parsed the majority of log lines, but were incapable of
distinguishing between log lines of the same category, e.g.\ rejecting a
mail delivery attempt.  The first parser reviewed~\cite{log-mail-analyser}
is the only previously published research in this area that the author is
aware of; their research aims to show that providing the data from log
files in a more accessible form is helpful to systems administrators.

The ten parsers have been reviewed and compared with \parsername{}, this
project's finished parser, to show how much effort would have been required
to use those to fulfil the aims and requirements of this project.  It is
important to compare and contrast newly developed algorithms and parsers
against those already available, to accurately judge what improvements, if
any, are delivered by the newcomers.

@@

Each of the reviewed parsers was tested with the \numberOFlogFILES{} test
log files described in \sectionref{parser efficiency}; \tableref{Summary of
parsers' features} summarises the results of this review.
